Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-06 article/article.Rmd
Commit 2024-09-06 19:12

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
# footnote:
#   - code: 1
#     text: "This is the first author footnote."
#   - code: 2
#     text: "Another author footnote."
# abstract: |
#   In this study we investigate the characteristics of global horizontal irradiance
#   enhancement events induced by clouds over Thessaloniki for the period 1994 -- 2023
#   using data recorded every one minute. We identified the cloud enhancement (CE)
#   events by creating an appropriate cloud-free irradiance reference with the use of a
#   radiative transfer model and aerosol optical depth data from a collocated Cimel sun
#   photometer and a Brewer spectrophotometer. We found a trend in CE events of
#   \(45.6\pm 21.9\,\text{cases}/\text{year}\), and a trend in the CE events
#   irradiation of \(116.9\pm 67.8\,\text{kJ}/\text{year}\). The peak of the CE events
#   was observed during May and June. The analysis of the total duration of CE events
#   showed that durations longer than 5 minutes are very rare, with exceptions lasting
#   over an hour and up to 140 minutes. Finally, we have detected enhancements above
#   the total solar irradiance at the top of the atmosphere of up to
#   \(400\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the cases below
#   \(200\,\text{W}/\text{m}^{2}\). The most active period of these extreme events is
#   spring -- early summer with a secondary peak in autumn.

@@

We studied the enhancement of the global horizontal irradiance induced by clouds over
Thessaloniki for the period 1994 -- 2023 using data recorded every one minute.  We
identified the cloud enhancement events (CE) by creating an appropriate cloud-free
irradiance reference with the use of a radiative transfer model and aerosol optical
depth data from a collocated Cimel sun photometer and a Brewer spectrophotometer. We
found a trend in CE events of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.N",slope],0))`\pm
`r round(2*yearlyCE[var=="GLB_diff.N",slope.sd],0)`\,\text{cases}/\text{year}$,
and a trend in the CE events irradiation of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.sum",slope],1))`\pm
`r round(2*yearlyCE[var=="GLB_diff.sum",slope.sd],1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June.  An analysis of the total
duration of CE events, showed that durations longer than 5 minutes are very rare,
with exceptions lasting over an hour and up to 140 minutes.  Finally, we have
detected enhancements above the total solar irradiance at the top of the atmosphere
for the same solar zenith angle of up to
$400\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$200\,\text{W}/\text{m}^{2}$. The most active season of these extreme events is
spring until early summer and autumn.

@@

<!-- Radiation data quality check -->
To further improve the quality of the irradiance data, a manual screening was
performed, to remove inconsistent and erroneous recordings that can occur
stochastically or systematically during the long operation of the instrument.  The
manual screening was aided by a radiation data quality assurance procedure, adjusted
for the site, which was based on the methods of Long and Shi\ [@Long2006;
@Long2008a].  Thus, problematic recordings have been excluded from further
processing.  Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by nearby
buildings, we have excluded all measurements with solar zenith angle (SZA) greater
than $`r 90 - BIO_ELEVA`^\circ$.  Finally, images from a sky camera have been used in
the manual inspection of the CE identification. The sky camera operates since 2012
and stores images in 5 min time steps.
<!--
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality of the
radiation measurements used in this study.
-->
<!-- Data selection -->
<!--
To preserve an unbiased representation of the data we applied a constraint, similar
the one used by @CastillejoCuberos2020. For each valid hour of day, there must exist
at least 45 minutes of valid measurements, including nighttime, when near sunrise and
sunset. Days with less than 5 valid hours were rejected completely.
-->
<!--
# CastillejoCuberos2020 There must be at least 5 h of valid data (Tests 1–7),
otherwise all data for this day is flagged as invalid. There must be at least This
work valid data 45 min of valid data (Tests 1–7) for an hour to be flagged as valid,
otherwise all minute data for this hour is flagged as invalid.
-->
An overview of the data analyzed for the study of the CE and ECE events is given in
Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily mean of the measured GHI.

```{r CLB-daily, echo=FALSE, fig.cap="Daily mean values of GHI in Thessaloniki for the period 1994 -- 2023."}
knitr::include_graphics("../images/P_daily_trend-3.png")
```

@@

Next we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrences normalized with the highest median
value, which occurs in June, depicting a clear seasonal cycle.  Although CE events
are present throughout the year, the most active months are May and June. During the
winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months, the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence, the
seasonally varying relative position of the sun, as well as the local landscape
characteristics that may influence the formation of the clouds.  Unfortunately, lack
of detailed data on cloud formation, type and location is not allowing further
analysis. The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.

>>>>

author:
  - name:        Athanasios N. Natsis
    email:       natsisphysicist@gmail.com
    affiliation: LAP
    correspondingauthor: true
    # footnote: 1
  - name:        Alkiviadis Bais
    email:       abais@auth.gr
    affiliation: LAP
  - name:        Charikleia Meleti
    email:       meleti@auth.gr
    affiliation: LAP
    # footnote: 1
address:
  - code:         LAP
    organization: Laboratory of Atmospheric Physics, Physics Department, Aristotle University of Thessaloniki
    # addressline:  1 main street
    city:         Thessaloniki
    postcode:     54124
    country:      Greece
  # - code:         Another University
  #   organization: Department
  #   addressline:  A street 29
  #   postcode:     2054 NX
  #   city:         Manchester,
  #   state:        State
  #   country:      The Netherlands
# footnote:
#   - code: 1
#     text: "This is the first author footnote."
#   - code: 2
#     text: "Another author footnote."
# abstract: |
#   In this study we investigate the characteristics of global horizontal irradiance
#   \(200\,\text{W}/\text{m}^{2}\). The most active period of these extreme events is
#   spring -- early summer with a secondary peak in autumn.

@@

In this study, we investigate the characteristics of global horizontal irradiance
enhancement events induced by clouds over Thessaloniki for the period 1994 -- 2023
using data recorded every one minute.  We identified the cloud enhancement (CE)
events by creating an appropriate cloud-free irradiance reference using a radiative
transfer model and aerosol optical depth data from a collocated Cimel sun photometer
and a Brewer spectrophotometer. We found a trend in CE events of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.N",slope],0))`\pm
`r round(2*yearlyCE[var=="GLB_diff.N",slope.sd],0)`\,\text{cases}/\text{year}$,
and a trend in the corresponding irradiation of
$`r plus_sign(round(yearlyCE[var=="GLB_diff.sum",slope],1))`\pm
`r round(2*yearlyCE[var=="GLB_diff.sum",slope.sd],1)`\,\text{kJ}/\text{year}$.
The peak of the CE events was observed during May and June.  An analysis of the total
duration of CE events, showed that with durations longer than 5 minutes are very rare,
with exceptions lasting over an hour and up to 140 minutes.  Finally, we have
detected enhancements above the total solar irradiance at the top of the atmosphere
for the same solar zenith angle of up to
$400\,\text{W}/\text{m}^{2}$, with the $75\,\%$ of the cases below
$200\,\text{W}/\text{m}^{2}$. Most of these  extreme events occur in
spring -- early summer, with a secondary peak in autumn.

@@

<!-- Radiation data quality check -->
To further improve the quality of the irradiance data, a manual screening was
performed, to remove inconsistent and erroneous recordings that can occur
stochastically or systematically during the long operation of the instrument.  The
manual screening was aided by a radiation data quality assurance procedure, adjusted
for the site, which was based on the methods of Long and Shi\ [@Long2006;
@Long2008a].  Thus, problematic recordings have been excluded from further
processing.  Furthermore, due to the significant measurement uncertainty in GHI when
the Sun is near the horizon, and due to some systematic obstructions by nearby
buildings, we have excluded all measurements with solar zenith angle (SZA) greater
than $`r 90 - BIO_ELEVA`^\circ$.  Finally, images from a sky camera have been used in
the manual inspection of the CE identification. The sky camera has been operating
since 2012 and stores images in 5 min time steps.  An overview of the GHI data used
in this study is given in
Figure\nobreakspace{}\ref{fig:CLB-daily}, as the daily means.

```{r CLB-daily, echo=FALSE, fig.cap="Timeseries of daily mean GHI measured at Thessaloniki for the period 1994 -- 2023."}
knitr::include_graphics("../images/P_daily_trend-3.png")
```

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the monthly box and
whisker plot of the CE number of occurrences normalized with the highest median
value, which occurs in June, depicting a clear seasonal cycle.  Although CE events
are present throughout the year, the most active months are May and June. During the
winter (December -- February), the number of CE cases is about $25\,\%$ of the
maximum, while in the intermediate months, the number of occurrences gradually ramps
between the maximum and minimum.  This seasonality is a combined effect of different
factors, among them the types of clouds, their frequency of occurrence, the
seasonally varying relative position of the sun, as well as the local landscape
characteristics that may influence the formation of the clouds.  Unfortunately, lack
of detailed data on cloud formation, type and location is not allowing further
analysis. The interannual variability of the monthly CE events is quite high as
manifested by the large monthly extremes, especially in the summer.
